---
title: "Cyclistic_data"
author: "Chigoziem"
date: "2023-03-11"
output: html_document
---

load the library
```{r} 

library(tidyverse)
library(skimr)
library(data.table)
library(lubridate)
library(hms)
library(ggplot2)

```
load the datasets
```{r}

jan01_df <- read_csv('202201_divvy_tripdata.csv')
feb02_df <- read_csv('202202_divvy_tripdata.csv')
mar03_df <-read_csv('202203_divvy_tripdata.csv')
apr04_df <- read_csv('202204_divvy_tripdata.csv')
may05_df <- read_csv('202205_divvy_tripdata.csv')
jun06_df <-read_csv('202206_divvy_tripdata.csv')
jul07_df <- read_csv('202207_divvy_tripdata.csv')
aug08_df <- read_csv('202208_divvy_tripdata.csv')
sep09_df <- read_csv('202209_divvy_tripdata.csv')
oct10_df <-read_csv('202210_divvy_tripdata.csv')
nov11_df <-read_csv('202211_divvy-tripdata.csv')
dec12_df <- read_csv('202212_divvy_tripdata.csv')
```

merge all  the datasets into one 
```{r}

cyclistic_df <- rbind(jan01_df,feb02_df,mar03_df,apr04_df,may05_df,jun06_df,jul07_df,aug08_df,sep09_df,oct10_df,nov11_df,dec12_df)
```

Remove individual data sets
```{r}
remove(jan01_df,feb02_df,mar03_df,apr04_df,may05_df,jun06_df,jul07_df,aug08_df,sep09_df,oct10_df,nov11_df,dec12_df)
```

create new data frame to contain new colums
```{r}
yearlistic_date_df <- cyclistic_df
```
calculate the ride_length by subtracting started_at ended_at and convert it to time in minutes.
```{r}
yearlistic_date_df$ride_length <- difftime(yearlistic_date_df$ended_at, yearlistic_date_df$started_at, units = 'mins')
```
create colum for day_of_week
```{r}
yearlistic_date_df$date <- as.Date(yearlistic_date_df$started_at)
yearlistic_date_df$day_of_week <- wday(yearlistic_date_df$started_at)
yearlistic_date_df$day_of_week <-format(as.Date(yearlistic_date_df$date), '%A')
yearlistic_date_df$month <-format(as.Date(yearlistic_date_df$date), '%m')
yearlistic_date_df$year <-format(as.Date(yearlistic_date_df$date), '%y')
yearlistic_date_df$day <- format(as.Date(yearlistic_date_df$date), '%d')
yearlistic_date_df$time <-format(as.Date(yearlistic_date_df$date),'%H:%M:%S')
yearlistic_date_df$time <-as_hms((yearlistic_date_df$started_at))
yearlistic_date_df$hour <-hour(yearlistic_date_df$time)
yearlistic_date_df<-yearlistic_date_df%>%mutate(season = 
                                            case_when(month=='01'~ 'Winter',
                                                      month == '02'~'Winter',
                                                      month == '03'~'Spring',
                                                      month == '04'~'Spring',
                                                      month == '05'~'Spring',
                                                      month == '06'~'Summer',
                                                      month =='07'~'Summer',
                                                      month =='08'~'Summer',
                                                      month == '09'~'Fall',
                                                      month == '10'~'Fall',
                                                      month == '11'~ 'Fall',
                                                      month == '12'~ 'Winter')
    )
```
create colum of hour of the day
```{r}
yearlistic_date_df<-yearlistic_date_df%>% mutate(time_of_day = 
                                             case_when(hour == '0' ~ 'Night',
                                                       hour =='1'~ 'Night',
                                                       hour == '2' ~ 'Night',
                                                       hour == '3' ~ 'Night',
                                                       hour == '4' ~ 'Night',
                                                       hour == '5' ~ 'Night',
                                                       hour == '6'  ~ 'Morning',
                                                       hour == '7' ~ 'Morning',
                                                       hour == '8' ~ 'Morning',
                                                       hour == '9' ~ 'Morning',
                                                       hour == '10' ~ 'Morning',
                                                       hour == '11' ~ 'Morning',
                                                       hour == ' 12' ~ 'Afternoon',
                                                       hour == '13' ~ 'Afternoon',
                                                       hour == '14' ~'Afternoon',
                                                       hour == '15' ~ 'Afternoon',
                                                       hour == '16'~ 'Afternoon',
                                                       hour == '17' ~ 'Afternoon',
                                                       hour == '18'~ 'Evening',
                                                       hour == '19' ~'Evening',
                                                       hour == '20' ~ 'Evening',
                                                       hour == '21' ~'Evening',
                                                       hour == '22' ~ 'Evening',
                                                       hour == '23' ~ 'Evening')
)
```
clean the dataset
```{r}
yearlistic_date_df<- distinct(yearlistic_date_df)
```
remove ride length that is than =  0 or less  
```{r}
yearlistic_date_df <- yearlistic_date_df[!(yearlistic_date_df$ride_length <=0),]
```
remove colums that is not needed
```{r}
yearlistic_date_df <- yearlistic_date_df %>%
  select(-c(ride_id,start_station_id,end_station_id,start_lat,end_lat,end_lng,start_lng))
```
view the final data
```{r}
view(yearlistic_date_df)
```
view the number of rows in the 
```{r}
nrow(yearlistic_date_df)
```
sum the ride length duplicated data
```{r}
sum(duplicated(yearlistic_date_df$ride_length))
```
view the head of the table
```{r}
head(yearlistic_date_df)
```
summarize the dataset
```{r}
summarize(yearlistic_date_df)
```

count the different member type


```{r}
yearlistic_date_df%>%   #count the different member type
  group_by(member_casual)%>%
  count(member_casual)
yearlistic_date_df%>%
  group_by(member_casual)%>%
  summarise(n=n())%>%
  mutate(percent = n*100/sum(n))
```
plot the member casual graph
```{r}
ggplot(data = yearlistic_date_df,mapping = aes
(x= member_casual)) +geom_bar()+labs(title = 'Members Against Casual')
```
calculate the different and total rides by member type
```{r}
yearlistic_date_df %>%
  group_by(member_casual,rideable_type)%>%
  count(rideable_type)
yearlistic_date_df%>%
  group_by(member_casual,rideable_type)%>%
  summarise(n=n())%>%
  mutate(percent = n*100/sum(n))
```

plot graph of different bike used by member types
```{r}
ggplot(data = yearlistic_date_df,mapping = aes
(x = rideable_type, fill = member_casual))+geom_bar()+labs(title = 'Different Bike Used')
```
total ride by member type in respect to hour of the day
```{r}
yearlistic_date_df %>%
  group_by(member_casual)%>%
  count(hour)%>%
  print(n=48)
#Total ride
yearlistic_date_df%>%
  count(hour)%>%
  print(n=24)
```
Total rides by member type (morning)
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(time_of_day == 'Morning') %>%
  count(time_of_day)
```
Total ride in the morning
```{r}
yearlistic_date_df %>%
  filter(time_of_day == 'Morning')%>%
  count(time_of_day)
```
Total rides by member type (Afternoon)
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(time_of_day == 'Afternoon')%>%
  count(time_of_day)

```

Total rides
```{r}
yearlistic_date_df %>%
  filter(time_of_day == 'Afternoon')%>%
  count(time_of_day)
```
total rides member type (evening)
```{r}
yearlistic_date_df %>%
  group_by(member_casual)%>%
  filter(time_of_day == 'Evening')%>%
  count(time_of_day)
```
total ride
```{r}
yearlistic_date_df%>%
  filter(time_of_day == 'Evening')%>%
  count(time_of_day)
```
total rides member type (night)
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(time_of_day == 'Night')%>%
  count(time_of_day)


yearlistic_date_df%>%
  filter(time_of_day == 'Night')%>%
  count(time_of_day)
```
total rides of member type(day of week)
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  count(day_of_week)
```
plot the graph of daily usage
```{r}

ggplot(data = yearlistic_date_df ,mapping =   aes
       (x = day_of_week,fill=member_casual))+geom_bar() + labs(title = 'Daily usage')
```

total rides of member type (day of month)
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  count(day)%>%
  print(n = 64)
#total rides
yearlistic_date_df%>%
  count(day)%>%
  print(n=31)
```
total monthly rides by member type
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  count(month )%>%
  print(n=24)
yearlistic_date_df%>%
  count(month)
```
plot the graph of monthly usage
```{r}
ggplot(data = yearlistic_date_df ,mapping =   aes
       (x = month,fill=member_casual))+geom_bar() + labs(title = 'Monthly usage')
```
total rides of member types by season
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(season == 'Spring')%>%
  count(season)
#total rides
yearlistic_date_df%>%
  filter(season == 'Spring')%>%
  count(seadon)
```
Total rides by member types in fall
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(season == 'Fall')%>%
  count(season)
```

Total rides by member types in summer 
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(season == 'Summer')%>%
  count(season)
```

Total rides by member types in Winter
```{r}
yearlistic_date_df%>%
  group_by(member_casual)%>%
  filter(season =='Winter')%>%
  count(season)
```

Total rides of member types  season
```{r}

yearlistic_date_df%>%
  group_by(season, member_casual)%>%
  count(season)
yearlistic_date_df%>%
  group_by(season)%>%
  count(season)
```
plot the graph of seasonal usage
```{r}
ggplot(data = yearlistic_date_df,mapping = aes
       (x= season , fill = member_casual))+geom_bar()+labs(title = 'Seasonal usage')

```

Ride length of member types
```{r}
yearlistic_avgRide <- mean(yearlistic_date_df$ride_length)
print(yearlistic_avgRide)
```
Ride length of member types
```{r}

yearlistic_avgRide <- mean(yearlistic_date_df$ride_length)
print(yearlistic_avgRide)
ggplot(data = yearlistic_date_df,mapping = aes
       (x = yearlistic_avgRide, fill = member_casual))+geom_bar()+labs(title = 'Average ride')
yearlistic_date_df%>%
  group_by(member_casual)%>%
  summarise_at(vars(ride_length),
               list(time = mean))
yearlistic_date_df%>%
  group_by(member_casual,rideable_type)%>%
  summarise_at(vars(ride_length),
               list(time = mean))
yearlistic_date_df%>%
  group_by(rideable_type)%>%
  summarise_at(vars(ride_length),
               list(time = mean))

yearlistic_date_df%>%
  group_by(member_casual,hour)%>%
  summarise_at(vars(ride_length),
               list(time = mean))%>%
  print(n =48)
yearlistic_date_df%>%
  group_by(member_casual,season)%>%
  summarise_at(vars(ride_length),
               list(time = mean))

 yearlistic_date_df%>%
  group_by(member_casual,rideable_type,season)%>%
  summarise_at(vars(ride_length),
               list(time = mean))
```

